---
title: "Simulation Challenge"
subtitle: "Generative Models and Monte Carlo Simulation"
format:
  html: default
execute:
  echo: false
  eval: true
---

# 🎲 Simulation Challenge - Monte Carlo Analysis

## Challenge Overview

**Your Mission:** Create a comprehensive Quarto document that simulates one or two investment strategies, analyzes the results, and demonstrates your ability to present counter-intuitive findings compellingly. Then render the document to HTML and deploy it via GitHub Pages from a new repository called "simulationChallenge."

# Complete Simulation

```{r}
#| label: complete-simulation

# Load required libraries
library(dplyr)
library(ggplot2)
library(purrr)
library(tibble)
library(scales)

# Set global options
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

# Game parameters
initial_balance <- 1000
heads_multiplier <- 1.5  # +50%
tails_multiplier <- 0.6  # -40%
years_to_retirement <- 35  # Assuming starting at age 20
num_simulations <- 10000

# Function to simulate a single game
simulate_game <- function(initial_balance, years, heads_mult, tails_mult) {
  balance <- initial_balance
  
  # Track the balance over time for this simulation
  balance_history <- numeric(years + 1)
  balance_history[1] <- balance
  
  for (year in 1:years) {
    # Flip coin (0 = tails, 1 = heads)
    coin_flip <- rbinom(1, 1, 0.5)
    
    if (coin_flip == 1) {
      # Heads: increase by 50%
      balance <- balance * heads_mult
    } else {
      # Tails: decrease by 40%
      balance <- balance * tails_mult
    }
    
    balance_history[year + 1] <- balance
  }
  
  return(list(
    final_balance = balance,
    balance_history = balance_history
  ))
}

# Run many simulations
set.seed(123)  # For reproducibility

# Run simulations
simulation_results <- map(1:num_simulations, function(i) {
  simulate_game(initial_balance, years_to_retirement, heads_multiplier, tails_multiplier)
})

# Extract final balances
final_balances <- map_dbl(simulation_results, ~ .x$final_balance)

# Create results dataframe
results_df <- tibble(
  simulation = 1:num_simulations,
  final_balance = final_balances,
  return_pct = (final_balances / initial_balance - 1) * 100
)

# Summary statistics
summary_stats <- results_df %>%
  summarise(
    mean_balance = mean(final_balance),
    median_balance = median(final_balance),
    min_balance = min(final_balance),
    max_balance = max(final_balance),
    mean_return = mean(return_pct),
    median_return = median(return_pct),
    prob_positive = mean(return_pct > 0),
    prob_break_even = mean(final_balance >= initial_balance)
  )

# Mathematical analysis
expected_annual_return <- 0.5 * 1.5 + 0.5 * 0.6  # 0.5 * 1.5 + 0.5 * 0.6 = 1.05
expected_total_return <- expected_annual_return^years_to_retirement
expected_final_balance <- initial_balance * expected_total_return

# Volatility analysis
volatility_stats <- results_df %>%
  summarise(
    std_dev = sd(return_pct),
    q25 = quantile(return_pct, 0.25),
    q75 = quantile(return_pct, 0.75),
    q90 = quantile(return_pct, 0.90),
    q95 = quantile(return_pct, 0.95),
    q99 = quantile(return_pct, 0.99)
  )

# Print results
cat("=== GAME PARAMETERS ===")
cat("\nInitial Balance: $", format(initial_balance, big.mark = ","))
cat("\nHeads: +50% (multiplier =", heads_multiplier, ")")
cat("\nTails: -40% (multiplier =", tails_multiplier, ")")
cat("\nYears to retirement:", years_to_retirement)
cat("\nNumber of simulations:", format(num_simulations, big.mark = ","))

cat("\n\n=== SIMULATION RESULTS ===")
cat("\nMean final balance: $", format(round(summary_stats$mean_balance, 2), big.mark = ","))
cat("\nMedian final balance: $", format(round(summary_stats$median_balance, 2), big.mark = ","))
cat("\nMin final balance: $", format(round(summary_stats$min_balance, 2), big.mark = ","))
cat("\nMax final balance: $", format(round(summary_stats$max_balance, 2), big.mark = ","))
cat("\nMean return: ", round(summary_stats$mean_return, 1), "%")
cat("\nMedian return: ", round(summary_stats$median_return, 1), "%")
cat("\nProbability of positive return: ", round(summary_stats$prob_positive * 100, 1), "%")
cat("\nProbability of breaking even or better: ", round(summary_stats$prob_break_even * 100, 1), "%")

cat("\n\n=== MATHEMATICAL ANALYSIS ===")
cat("\nExpected annual multiplier:", round(expected_annual_return, 4))
cat("\nExpected total multiplier over", years_to_retirement, "years:", round(expected_total_return, 2))
cat("\nExpected final balance: $", format(round(expected_final_balance, 2), big.mark = ","))
cat("\nExpected return: ", round((expected_final_balance / initial_balance - 1) * 100, 1), "%")

cat("\n\n=== VOLATILITY ANALYSIS ===")
cat("\nStandard deviation of returns:", round(volatility_stats$std_dev, 1), "%")
cat("\n25th percentile:", round(volatility_stats$q25, 1), "%")
cat("\n75th percentile:", round(volatility_stats$q75, 1), "%")
cat("\n90th percentile:", round(volatility_stats$q90, 1), "%")
cat("\n95th percentile:", round(volatility_stats$q95, 1), "%")
cat("\n99th percentile:", round(volatility_stats$q99, 1), "%")
```

# Visualizations

```{r}
#| label: visualizations

# Distribution of Final Balances
p1 <- ggplot(results_df, aes(x = final_balance)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7, color = "black") +
  geom_vline(xintercept = initial_balance, color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = summary_stats$mean_balance, color = "green", linetype = "dashed", size = 1) +
  geom_vline(xintercept = summary_stats$median_balance, color = "orange", linetype = "dashed", size = 1) +
  scale_x_continuous(labels = dollar_format(), trans = "log10") +
  labs(
    title = "Distribution of Final Balances (Log Scale)",
    subtitle = "Red line = Initial $1,000 | Green line = Mean | Orange line = Median",
    x = "Final Balance ($)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )

# Distribution of Returns
p2 <- ggplot(results_df, aes(x = return_pct)) +
  geom_histogram(bins = 50, fill = "darkgreen", alpha = 0.7, color = "black") +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = summary_stats$mean_return, color = "blue", linetype = "dashed", size = 1) +
  geom_vline(xintercept = summary_stats$median_return, color = "orange", linetype = "dashed", size = 1) +
  labs(
    title = "Distribution of Percentage Returns",
    subtitle = "Red line = Break-even (0%) | Blue line = Mean | Orange line = Median",
    x = "Return (%)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )

# Sample Paths Over Time
sample_indices <- sample(1:num_simulations, 100)
sample_histories <- map(sample_indices, ~ simulation_results[[.x]]$balance_history)

years <- 0:years_to_retirement
path_data <- map_dfr(1:length(sample_histories), function(i) {
  tibble(
    simulation = i,
    year = years,
    balance = sample_histories[[i]]
  )
})

p3 <- ggplot(path_data, aes(x = year, y = balance, group = simulation)) +
  geom_line(alpha = 0.3, color = "steelblue") +
  scale_y_continuous(labels = dollar_format(), trans = "log10") +
  labs(
    title = "Sample of 100 Simulation Paths",
    subtitle = "Each line represents one possible outcome over 35 years",
    x = "Year",
    y = "Account Balance ($, Log Scale)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12)
  )

# Display plots
print(p1)
print(p2)
print(p3)
```
